---
title: 'COllege Major Outlook: Difference between STEM and non-STEM - Spring 2020'
author: 
  - name: 'Sathwik Kesappragda'
email: ''
output: rmarkdown::github_document
 # html_document:
 #   fig_width: 10 
 #   fig_height: 12 
 # toc: yes
toc_depth: 4
---
# **PROJECT WORK**

Project Proposal Questions// Brainstorming ~ pertaining to my portion only   
Salary difference between STEM majors and NON-STEM majors   
How do different categories of majors stack up against each other like engineering majors to science to math to arts   
What majors are the most popular college degrees  
Unemployment rates respective to the major   


**loading necessary packages**

```{r, collapse = T}
#libraries used for project
library(tidyverse) # for `ggplot2`, `dplyr`, and more
library(readr) #read excel csv files, downloaded from kaggle 
library(dplyr)
library(gridExtra)
library(ISLR)
library(tibble)
library(boot)
library(scales)
library(class)
```

### (I.) Data Wrangling  & Cleaning

**Data was extracted from Kaggle website, using readr library to read and store the csv file/our data of interest into 'recentgrads' object.**

```{r, collapse = T}
##main data
recentgrads <- read_csv(file = "C:/Users/sathw/Downloads/DS/collegemajor_project/raw/recent-grads.csv")
recentgrads

### data we thought of using
# allages <- read.csv(file = "all-ages.csv") 
# gradstudents <- read.csv(file = "grad-students.csv")
# allages
# gradstudents
```


**Since my portion of the project focuses on drawing comparisons between STEM and NONSTEM, I needed to acquire the proper subsets. Furthermore, I chose which category falls under which sector.**         
**There are about 16 Major categories. 86 majors in STEM and 78 majors in NONSTEM. Pyschology appears in both categories.**    

```{r}
## go through each of the files and determine whether each record is stem or non-stem
## recentgrads data 
majors <- recentgrads %>%
  select(Major_category) %>%
  distinct()
majors

STEM <- filter(recentgrads, recentgrads$Major_category == "Engineering" 
               | recentgrads$Major_category == "Physical Sciences" 
               | recentgrads$Major_category == "Computers & Mathematics"
               | recentgrads$Major_category == "Agriculture & Natural Resources"
               | recentgrads$Major_category == "Health"
               | recentgrads$Major_category == "Biology & Life Science")

STEM

NONSTEM <- filter(recentgrads, recentgrads$Major_category == "Business"
                  | recentgrads$Major_category == "Law & Public Policy"
                  | recentgrads$Major_category == "Industrial Arts & Consumer Services"
                  | recentgrads$Major_category == "Arts"
                  | recentgrads$Major_category == "Social Science" 
                  | recentgrads$Major_category == "Education"
                  | recentgrads$Major_category == "Humanities & Liberal Arts"
                  | recentgrads$Major_category == "Pyschology & Social Work"
                  | recentgrads$Major_category == "Communications & Journalism"
                  | recentgrads$Major_category == "Interdisciplinary")

NONSTEM
```

### (II.) Assessing Data  

**Briefing over some descriptive statistics I wanted to analyze.**  
**Petroleum Engineering has low unemployment rate and a high median salary. Biology, Computer Science, Business, and Accounting are the most impacted majors. There exists a major called Metallurgical engineering which is the study of metals.**

```{r, collapse = T}
#finding top 5 stem median salaries
STEM_mediansalary <- group_by(STEM, Major)
sal_sum <- summarise(STEM_mediansalary, med_salary = Median)
med_sal_STEM <- arrange(sal_sum, desc(med_salary))
head(med_sal_STEM, 5)

#finding top 5 non stem median salaries
NONSTEM_mediansalary <- group_by(NONSTEM, Major)
nsal_sum <- summarise(NONSTEM_mediansalary, med_salary = Median)
med_sal_NONSTEM <- arrange(nsal_sum, desc(med_salary))
head(med_sal_NONSTEM, 5)

#finding top5 most popular in STEM
STEM_majors <- group_by(STEM, Major)
overview <- summarise(STEM_majors, total = sum(Total))
most_popularSTEM <- arrange(overview, desc(total))
head(most_popularSTEM, 5)

#finding top5 most popular in NONSTEM
NONSTEM_majors <- group_by(NONSTEM, Major)
overview <- summarise(NONSTEM_majors, total = sum(Total))
most_popularNONSTEM <- arrange(overview, desc(total))
head(most_popularNONSTEM, 5)

#finding top 5 lowest unemployment rates for STEM 
unemploymentrateSTEM_majors <- group_by(STEM, Major)
rate <- summarise(unemploymentrateSTEM_majors, unemployment_rate = Unemployment_rate)
least_unemployment_rate_STEM <- arrange(rate, -desc(unemployment_rate))
head(least_unemployment_rate_STEM, 5)


#finding top 5 lowest unemployment rates  for NONSTEM
unemploymentrateNONSTEM_majors <- group_by(NONSTEM, Major)
un_rate <- summarise(unemploymentrateNONSTEM_majors, unemployment_rate = Unemployment_rate)
least_unemployment_rate_NONSTEM <- arrange(un_rate, -desc(unemployment_rate))
head(least_unemployment_rate_NONSTEM, 5)
```

### (III.) Exploratory Data Analysis (EDA)  

**Data visualization**
```{r, collapse = T}
## Objective: Show which major classes fall under which sector; the biggest categories that make up each division
par(mar = c(7, 7, 7, 7))
#coxcomb for STEM by category
STEMbar <- ggplot(data = STEM) +
  geom_bar(mapping = aes(x = Major_category, fill = Major_category), show.legend = F, width = 1) + 
  theme(aspect.ratio = 1) +
  labs(x = NULL,  y = NULL) 
plot2 <- STEMbar + coord_polar()
grid.arrange(plot2, ncol = 2)

#coxcomb for NONSTEM by category
NONSTEMbar <- ggplot(data = NONSTEM) +
  geom_bar(mapping = aes(x = Major_category, fill = Major_category), show.legend = F, width = 1) + 
  theme(aspect.ratio = 1) +
  labs(x = NULL,  y = NULL) 
plot2 <- NONSTEMbar + coord_polar()
grid.arrange(plot2, ncol = 1, nrow =1 )
```

```{r, collapse = T}
#histogram for Total number of college students by Major Category for both STEM & NONSTEM 
#STEM
ggplot(data = STEM, mapping = aes(x = Major_category, y = Total)) + 
  geom_histogram(mapping = aes(fill = Major_category), stat = "identity", binwidth = 1000, position = "dodge") + 
  coord_flip()
#NONSTEM
ggplot(data = NONSTEM, mapping = aes(x = Major_category, y = Total)) + 
  geom_histogram(mapping = aes(fill = Major_category), stat = "identity", binwidth = 1000, position = "dodge") + 
  coord_flip()
```

```{r, collapse = T}
#histogram for top5 stem median salaries
top5stem <-  head(med_sal_STEM, 5)
ggplot(data = top5stem, mapping= aes(x = Major, y = med_salary)) +
  geom_histogram(mapping = aes(fill= Major),  stat = "identity", show.legend = F) +
  scale_y_continuous(labels = scales::comma) +
  geom_text(mapping = aes(label = med_salary), vjust = 1) +
  labs(title = "Top 5 Median Salaries STEM Major", x = "Major Category", y= "Median Salary") +
  theme(axis.text.x = element_text(size = 10,angle = 55,hjust = 1,vjust = 1))

#histogram for top5 nonstem median salaries
top5nonstem <- head(med_sal_NONSTEM, 5)
ggplot(data = top5nonstem, mapping= aes(x = Major, y = med_salary)) +
  geom_histogram(mapping = aes(fill= Major),  stat = "identity", show.legend = F) +
  scale_y_continuous(labels = scales::comma) +
  geom_text(mapping = aes(label = med_salary), vjust = 1) +
  labs(title = "Top 5 Median Salaries Non-STEM Major", x = "Major Category", y= "Median Salary") +
  theme(axis.text.x = element_text(size = 10,angle = 55,hjust = 1,vjust = 1))
```

```{r, collapse = T}
##boxplots to compare the number of college jobs for each specific major division 
ggplot(data = STEM) +
geom_boxplot(mapping = aes(x = Major_category, y = College_jobs, color = Major_category), show.legend = F) +
  coord_flip()

ggplot(data = STEM, mapping = aes(x = Major_category, y = College_jobs)) +
  geom_violin() +
  coord_flip()

#Frequency plots to show which industry has the most amount of people employed: STEM or NONSTEM
ggplot(data = STEM) +
  geom_freqpoly(mapping = aes(x = Employed, color = Major_category),
                binwidth = 100000) + 
  scale_x_continuous(labels = scales::comma) +
  labs(title = "STEM Number of Employed")

ggplot(data = NONSTEM) +
  geom_freqpoly(mapping = aes(x = Employed, color = Major_category),
                binwidth = 100000) +
  scale_x_continuous(labels = scales::comma) +
  labs(title = "Non-STEM Number of Employed")
```

### (IV.) Modeling  

```{r, collapse = T}
#make a new column; binary => 0s and 1s to label which major is STEM or NONSTEM
recentgrads_SorN <- recentgrads %>% #Creating new table define STEM and NONSTEM as 0s and 1s
  group_by(Major, Major_category) %>%
  mutate(SorN = if(Major_category == "Engineering" | Major_category == "Physical Sciences" | 
                   Major_category == "Computers & Mathematics" | Major_category == "Agriculture & Natural Resources"| Major_category == "Health"| 
                   Major_category == "Biology & Life Science"){
    n = 1 #STEM is 1 
  }else{
    n = 0 #NONSTEM is 0
  }, default = if(Major_category == "Engineering" | Major_category == "Physical Sciences" | 
                   Major_category == "Computers & Mathematics" | Major_category == "Agriculture & Natural Resources"| Major_category == "Health"| 
                   Major_category == "Biology & Life Science"){
    n = "Yes" #STEM is Yes 
  }else{
    n = "No" #NONSTEM is No
  }) %>%
  select(Major, Major_category, SorN, default, Total, Employed, Unemployed, Unemployment_rate, Median, College_jobs)
recentgrads_SorN
```

**Ran a logistic regression between the 'SorN' versus 'Total', 'Unemployment_rate', 'Median', and 'College_jobs'.**    
**Suprisingly for college_jobs and unemployment_rate the p-values are greater than alpha showing that they are NOT significant predictors for influencing one's choice to join either STEM or NONSTEM. On the other hand, median salary and total number of students have p-values less than alpha = 0.10 indicating that they're contributors to the decision between STEM and NONSTEM.**  
```{r, collapse = T}
# set the random seed so that the analysis is reproducible
set.seed(168)
recentgrads_SorN_idx = sample(nrow(recentgrads_SorN), 86) #splitting the data in half
recentgrads_SorN_trn = recentgrads_SorN[recentgrads_SorN_idx, ] #training data
recentgrads_SorN_tst = recentgrads_SorN[-recentgrads_SorN_idx, ] #testing data
rcg_trn_lm = recentgrads_SorN_trn
rcg_tst_lm = recentgrads_SorN_tst


#logit.fit.all
model_glm_all = glm(SorN ~ Total+Unemployment_rate+Median+College_jobs , 
                    data = recentgrads_SorN_trn, 
                    family = "binomial")
summary(model_glm_all)
```


```{r, collapse = T}
model_glm = glm(SorN ~ Median, data = recentgrads_SorN_trn, family = "binomial")

plot(SorN ~ Median, data = rcg_trn_lm, 
     col = "darkorange", pch = "|", ylim = c(0.0, 1.0), xlim=c(20000, 120000), 
     main = "Using Logistic Regression for Classification") 
#axis(1, at = seq(20000, 120000, by = 10000), las=2)
abline(h = 0, lty = 3)
abline(h = 1, lty = 3)
abline(h = 0.5, lty = 2)
curve(predict(model_glm, data.frame(Median = x), type = "response"), 
      add = TRUE, lwd = 3, col = "dodgerblue")
abline(v = -coef(model_glm)[1] / coef(model_glm)[2], lwd = 2)

  
#Estimated conditional probability curve
ggplot(data = recentgrads_SorN, mapping = aes(x = Median)) + ylab("StemorNonSTEM") +
  geom_histogram(mapping = aes(fill = default), 
                 binwidth = 25000, position = 'fill') + 
  geom_smooth(aes(y = SorN), 
              method="glm", method.args=list(family="binomial"), se = F, size = 3) +
  labs(title = "Median Salary: STEMvNONSTEM", x = "Median Salary", y = "STEM(1)NONSTEM(0)")

#scatterplot to visualize Total number of students in that major vs Median salary; group_by STEM or NONSTEM
ggplot(recentgrads_SorN) + 
  geom_point(mapping = aes(x = Total, y = Median, colour = default)) +
  scale_y_continuous(labels = scales::comma) +
  scale_x_continuous(labels = scales::comma)



ggplot(data = recentgrads_SorN, mapping = aes(x = Median, y = SorN)) + 
  geom_point(mapping = aes(color = default)) +
  scale_x_continuous(labels = scales::comma)


ggplot(data = recentgrads_SorN, mapping = aes(x = Median)) + 
  geom_histogram(mapping = aes(fill = default), binwidth = 20000, position = "dodge")

gg <- ggplot(recentgrads_SorN) + 
  geom_point(aes(x = Total, y = Median, color = default)) # plot test point 
x.test <- c(40000, 50000)
gg + geom_point(x = x.test[1], y = x.test[2], shape = 'X', size = 10) +
  scale_y_continuous(labels = scales::comma) +
  scale_x_continuous(labels = scales::comma)

```


**Nearest Centroid (NC) Classifier**  
**Calculates the difference in means of median salary between the sectors. Plotted a line segment to visualize the distance.** 
```{r, collapse = T}
# compute the observed class means 
(obs.means <- recentgrads_SorN %>% 
    group_by(default) %>% 
    select(default, Median, Total) %>%  
    summarise_all(mean, na.rm = TRUE) #ignore mean of total, only to help draw the line segment 
) 
(dist <- obs.means %>% mutate(dist = sqrt((Total - x.test[1])^2 + (Median- x.test[2])^2)))
(gg <- ggplot(recentgrads_SorN) + 
    geom_point( mapping = aes(x = Total, y = Median, color = default), alpha = 1.0) +
    geom_point(x = x.test[1], y = x.test[2], shape = 'X', size = 10) + 
    # plot the two class means 
    geom_point(data = obs.means, shape = 'X', size = 10, 
               aes(x = Total, y = Median, color = default)) +
  scale_y_continuous(labels = scales::comma) +
  scale_x_continuous(labels = scales::comma)+ 
    geom_segment(data = dist, aes(x = x.test[1], y = x.test[2], xend = Total, yend = Median))) #add line segment to graph

```

**Misclassification Rate = 30.00578% <- The perecentage of training and testing data thats is misclassified.**
```{r}
#logistic regression
logit.poly.fit <- glm(SorN ~ Median, family = binomial(), data = recentgrads_SorN) 
#predict the conditional prob
logit.fit.prob <- predict(logit.poly.fit, type = "response") 
#Bayes rule
logit.fit.class <- as.factor(ifelse(logit.fit.prob > 0.5, "Yes", "No"))
#Misclassification error rate
mean(recentgrads_SorN$default != logit.fit.class)
```










```{r}

#### SCRATCHWORK JUNK ### IGNORE


# #attempt to combine two tables by joins 
# recentgrads1 <- recentgrads %>%
#   select(-Rank, -ShareWomen, -College_jobs, -Non_college_jobs, -Low_wage_jobs)
# 
# recentgrads1 %>%
#   inner_join(gradstudents, by = c("Major", "Major_category")) %>%
#   select(-Grad_premium, -Grad_share, -Nongrad_total, -Nongrad_employed, -Nongrad_full_time_year_round, -Nongrad_unemployed,-Nongrad_unemployment_rate, -Nongrad_median, -Nongrad_P25, -Nongrad_P75)
# 
# recentgrads1


# training data
# X_recentgrads_SorN_trn = recentgrads_SorN_trn[, -1]
# y_recentgrads_SorN_trn = recentgrads_SorN_trn$SorN
# 
# # testing data
# X_recentgrads_SorN_tst = recentgrads_SorN_tst[, -1]
# y_recentgrads_SorN_tst = recentgrads_SorN_tst$SorN
# 
# 
# 
# 
# knn(train = recentgrads_SorN_trn, 
#          test  = recentgrads_SorN_tst, 
#          cl    = recentgrads_SorN_tst$SorN, 
#          k     = 2, prob = TRUE)



# studentsinSTEM <- function(STEM){
#   sumtotal = 0 
#   for(i in 1:ncol(STEM)){
#     for(i in 1:nrow(STEM)){
#       sumtotal = sumtotal + STEM$Total
#     }
#   }
#   return(sumtotal)
# }
#   
# studentsinSTEM

# total <- STEM %>%
#   summarize(sum = sum(STEM$Total))
# 
# total
#   

#visualization 
# gender_totals <- tibble(mentotal = STEM$Men, womentotal = STEM$Women)
# gg <- ggplot(data = STEM, mapping = aes(x = Total, y = Major_category))  
# gg + geom_histogram(mapping = aes(fill = Men), binwidth =10000, position = "dodge")
# ggplot(data=STEM, mapping = aes(x= Men, Women, y = Total, fill = Major_category)) +
#   geom_bar(stat = "identity", position = "dodge") +
#   coord_flip()
# ggplot(data = STEM, aes(x = Major_category, y= Total, group = Major_category)) +
#   geom_bar(aes(Men, Women, fill = Major_category), stat = "identity", position = "dodge") +
#   coord_flip()


# top5urSTEM <- head(least_unemployment_rate_STEM, 5)
# ggplot(data=top5urSTEM, mapping = aes(x= Major, y = unemployment_rate, color = Major)) +
#   geom_bar(stat = "identity") +
#   coord_flip()

# total_jobs <- STEM %>%
#   select(Major)
# gender_totals <- tibble(mentotal = STEM$Men, womentotal = STEM$Women)

# gg <- ggplot(data=STEM, mapping = aes(x=Men)) +
# theme(legend.position = "bottom")
# plot1 <- gg + geom_histogram(aes(fill=Major_category), binwidth=0.5)
# plot2 <- gg + geom_freqpoly(aes(color=Major_category), binwidth=0.5)
# grid.arrange(plot1, plot2, ncol=2)


# library(plotROC) # extract conditional prob. P(default="Yes" | balance,income,student) 
# logit.fit.prob <- predict(model_glm_all, type = "response") 
# roc.df <- tibble(observed = recentgrads_SorN$SorN, 
#                  predicted = logit.fit.prob) 
# ggplot(data = roc.df, mapping = aes(d = observed, m = predicted)) + 
#   geom_roc(labels=F)


# (confusion.matrix <- table(recentgrads_SorN$SorN, recentgrads_SorN$SorN)) 
# recentgrads_SorN.knn <- recentgrads_SorN
# recentgrads_SorN.knn$default.knn.pred <-
#   knn(X_recentgrads_SorN_trn, X_recentgrads_SorN_tst,
#     # train = select(recentgrads_SorN,Median, Total), # training data predictors [matrix/data.frame] 
#        cl, # training data class labels [vector]
#     #   test = select(recentgrads_SorN,Median, Total), # test data [matrix/data.frame] 
#       k = 5, prob = TRUE) # set K - # of nearest neighbors
# knn_misclassification_rate = mean(recentgrads_SorN$default !=recentgrads_SorN.knn$default.knn.pred)




# rcg_trn_lm$SorN = as.numeric(rcg_trn_lm$SorN) - 1
# rcg_tst_lm$SorN = as.numeric(rcg_tst_lm$SorN) - 1

# 
# model_glm = lm(SorN ~ Median, data = recentgrads_SorN_trn, data = "binomial" )
# 
# plot(SorN ~ Median, data = rcg_trn_lm, 
#      col = "darkorange", pch = "|", ylim = c(-0.2, 1),
#      main = "Using Logistic Regression for Classification")
# abline(h = 0, lty = 3)
# abline(h = 1, lty = 3)
# abline(h = 0.5, lty = 2)
# curve(predict(model_glm, data.frame(Median = x), type = "response"), 
#       add = TRUE, lwd = 3, col = "dodgerblue")
# abline(v = -coef(model_glm)[1] / coef(model_glm)[2], lwd = 2)


# recentgrads_SorN.knn <- recentgrads_SorN
# recentgrads_SorN$default.knn.pred <-
#   knn(train = select(recentgrads_SorN,Total,Median), # training data predictors [matrix/data.frame] 
#       cl = recentgrads_SorN$default, # training data class labels [vector] 
#       test = select(recentgrads_SorN,Total,Median), # test data [matrix/data.frame] 
#       k = 5) # set K - # of nearest neighbors
# recentgrads_SorN.knn 


# test_grid <- expand.grid(Median = seq(0, 120000, length = 1000), 
#                          Total = seq(0, 400000, length = 1000)) %>% 
#   as_tibble()

# mean.pos <- filter(obs.means, default == "Yes") %>% select(Total, Median) 
# mean.neg <- filter(obs.means, default == "No") %>% select(Total, Median)
# # make a grid of test points 
# test_grid <- expand.grid(Total = seq(0, 80000, length = 10000), Median = seq(0, 75000, length = 10000)) %>% as_tibble()
# # for each test point, calculate the Euclidean distances to class means 
# dist_pos <- base::apply(test_grid, 1, function(x) sqrt(sum((x - mean.pos)^2))) 
# dist_neg <- base::apply(test_grid, 1, function(x) sqrt(sum((x - mean.neg)^2))) 
# # add distance columns to the test_grid data frame 
# test_grid <- test_grid %>% add_column(dist_pos = dist_pos, dist_neg = dist_neg)
# # decide which class mean each test point is closest to 
# test_grid <- test_grid %>% mutate(y_pred = as.factor(ifelse(dist_pos<dist_neg, "Yes","No")))


```